---
title: "lab_08"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.


**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key

```{r echo=FALSE}
census_api_key("8d7ad09f8f8dd3b90a48c6afa09aa9e03b668fa5", install=TRUE)
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2019 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2019 5-year ACS and save them to a dataframe

```{r}
# Write code here to load the ACS variables
acs_data <- load_variables(2019, "acs5", cache = TRUE)

```

## Load PPP data

You'll also need to load an RDS file of PPP loan applications that were removed from the original data by lenders in July.

**Task** read the RDS file and save it to a dataframe

```{r}
#Write code here to read the RDS file
md_ppp_loans <- read_rds("data/lab_08.rds")
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2019 5-year estimates? You will need to find the variable for median age first.
**A1** Baltimore County 

```{r}
md19 <- get_acs(geography = "county",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2019)
view(md19)
```

**Q2** Which Maryland counties have a median age of at least 50?
**A2** Talbot County and Worcester County have a median age of at least 50. 

```{r}
md19 %>%
  filter(estimate > 50)

```

**Q3** Which Maryland counties have the highest percentage of 5-17 year-olds that speak Spanish at home? Start with the code you wrote for Q4 and add a second dataframe with the total population of 5-17 year-olds by county. For each dataframe, rename the `estimate` column to a meaningful name you've given the ACS variable, remove the `variable` columns in each using select and then join them based on GEOID and NAME. Then you can add a column to calculate the percentage of the total 5-17 population that speaks Spanish at home.
**A3** Prince George's County, Montgomery County, and Caroline County all have the highest percentages of 5-17 year-olds that speak Spanish at home. 

```{r}
md_speak_spanish <- get_acs(geography = "county",
              variables = c( percentage_spanish= "B16007_004"),
              state = "MD",
              year = 2019) %>%
rename(speak_span= estimate)

md_pop <- get_acs(geography = "county",
              variables = c(age = "B06001_003"),
              state = "MD",
              year = 2019)%>%
rename(teens= estimate)

md_pop %>%
  inner_join(md_speak_spanish, by=c("GEOID", "NAME")) %>%
  select(-variable.x, -variable.y)%>%
   mutate(percent= (speak_span/teens)*100) %>%
  arrange(desc(percent))
 
```

**Q4** Given the counties with the largest percentage of 5-17 year-olds that speak Spanish at home, what stands out to you about the data? How would you describe the top 5 counties? How would you describe Garrett County's numbers?
**A4** What stands out to me is that, in the top 5 counties, those with a higher population don't necessarily have the most kids that speak Spanish at home. For example, Frederick County has more kids (43949) than Caroline County (5832), but the latter still has a higher percentage of the spanish-speaking population at home (10.2). I would describe the top 5 counties as being very different from one another. They all have different numbers of kids as well as how many of them speak Spanish at home. The top two counties makes sense as Prince George's County and Montgomery County are the two biggest in Maryland. While Caroline County may be one of the smallest, it could be a hotspot for Spanish-speaking immigrants. I would describe Garrett County's numbers as being so low due to it being a more white county. Garrett County is also one of the smallest counties in Maryland, so it would have lower numbers. 

**Q5** Which 5-digit zip code had the most removed PPP applications, and what is the percentage of Black population in that zip code according to the ACS? To answer this question, you will need to creeate separate dataframes with ACS data on total population by zip code and Black population by zip code. In each dataframe, rename the `estimate` column and remove the `variable` column. You'll need to combine the two dataframes in order to calculate the percentage, and then you can combine it with the PPP data. Remember: the PPP data may not have only 5-digit zip codes!
**A5**

```{r}

#black population by zipcode dataframe
black_pop <- get_acs(geography="zcta", 
                       variables = c(population ="B02001_003"),
                        year = 2019) %>%
rename(total_black=estimate)%>%
  select(-variable)

#total population by zipcode dataframe
total_pop <- get_acs(geography="zcta", 
                       variables = c(population="B01003_001"), 
                        year = 2019)%>%
rename(total_pop=estimate)%>%
  select(-variable)

#joining data
total_with_black <- total_pop %>%
  inner_join(black_pop, by=c("GEOID")) %>%
  mutate(percent_black_pop= (total_black/total_pop)*100)

#ppp loans
removed_ppp_loans <- removed_ppp_loans %>%
mutate(zip5 = str_sub(zip, 1, 5))

data_with_zips <- removed_ppp_loans%>%
  inner_join(total_with_black, by=c(zip5="GEOID"))

data_with_zips %>%
group_by(zip5, percent_black_pop) %>%
  summarise(count_loans=n()) %>%
  arrange(desc(count_loans))
  

```

**Q6** How would you describe to an editor the top 10 zip codes with the most removed application in terms of Black population? Where are they located, and what do they have in common? Are there any zip codes that seem different, and if so why?
**A6** The top 10 zip codes have a significant amount of the Black population, with 75% to 95%. They all start with the same numbers (606) except for one (30349). The ones that start with 606 are all in Illinois while the 303 is located in Georgia. Both Georgia and Illinois are known for having a significant part of the Black population.  

**Q7** What other columns from the PPP data could be useful in reporting on what happened to this removed loans?
**A7** Other helpful columns in the ppp data are addresses and business type. These could provide insight on where loans are being taken from as well as if certain businesses got some removed. 
